A formal treatment of bidirectional typing

### 2023-10-12 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Revise until S6

\begin{theorem}[Generalised Mode Preprocessing]\label{thm:generalised-mode-decoration}
For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
\end{theorem}

@@

\begin{lemma}\label{thm:adjustment}
For any raw term $V |-_{\Sigma,\erase\Omega} t$, if
\begin{equation}\label{eq:strict-derivation}
V |-_{\Sigma,\Omega} t^{\dir{d'}\,g'\,\BTrue} \quad\text{for some mode~$\dir{d'}$ and boolean~$g'$}
\end{equation}
then for any mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
\end{lemma}

@@

\begin{proof}[\cref{thm:generalised-mode-decoration}]
By \cref{thm:adjustment}, it suffices to prove~\eqref{eq:strict-derivation} by induction on~$t$.

\begin{itemize}
\item $\Rule{Var}$ is mapped to $\SynRule{Var}$.
\item $\Rule{Anno}$:
Let $t = (t' \bbcolon A)$.
By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} {t'}^{\chk\,g\,s}$ for some booleans $g$~and~$s$, to which we apply $\SynRule{Anno}$.
\item $\Rule{Op}$:
Let $t = \tmOpts$.
By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} {t_i}^{\dir{d_i}\,g_i\,s_i}$ for each sub-term~$t_i$; apply $\Rule{Op}$ to all the derivations.
\end{itemize}
\end{proof}

@@

\begin{corollary}[Mode Preprocessing]\label{thm:mode-decoration}
  It is decidable whether\/ $V |-_{\Sigma,\Omega} \isTerm{t}^\dir{d}$.%for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$
\end{corollary}

@@

\begin{proof}[Proof of \cref{thm:annotatability}]
By \cref{thm:generalised-mode-decoration}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.
Perform induction on this generalised mode derivation to construct a bidirectional typing derivation in the same mode (and also the term~$t'$ and a proof of $t' \sqsupseteq t$, which are determined by the derivations and omitted here).
\begin{itemize}
\item $\SynRule{Var}$, $\SynRule{Anno}$, and $\Rule{Op}$:
The outermost rule of the given typing derivation must be the corresponding one; apply the corresponding bidirectional typing rule to the induction hypotheses.
\item $\ChkRule{Sub}$:
Apply $\ChkRule{Sub}$ to the induction hypothesis.
\item $\SynRule{Missing}$:
This is the interesting case where we map a $\SynRule{Missing}$ rule to an $\SynRule{Anno}$ rule and add to the term a type annotation, which comes from the given typing derivation.
\end{itemize}
\end{proof}

>>>>

\begin{theorem}[Generalised mode preprocessing]\label{thm:generalised-mode-decoration}
For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some $g$~and~$s$.
\end{theorem}

@@

\begin{lemma}\label{thm:adjustment}
For any raw term $V |-_{\Sigma,\erase\Omega} t$, if
\begin{equation}\label{eq:strict-derivation}
V |-_{\Sigma,\Omega} t^{\dir{d'}\,g'\,\BTrue} \quad\text{for some mode~$\dir{d'}$ and boolean~$g'$}
\end{equation}
then for any mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some $g$~and~$s$.
\end{lemma}

@@

\begin{proof}[of \cref{thm:generalised-mode-decoration}]
By \cref{thm:adjustment}, it suffices to prove~\eqref{eq:strict-derivation} by induction on~$t$.

\begin{itemize}
\item $\Rule{Var}$ is mapped to $\SynRule{Var}$.
\item $\Rule{Anno}$:
Let $t = (t' \bbcolon A)$.
By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} {t'}^{\chk\,g\,s}$ for some $g$~and~$s$, to which we apply $\SynRule{Anno}$.
\item $\Rule{Op}$:
Let $t = \tmOpts$.
By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} {t_i}^{\dir{d_i}\,g_i\,s_i}$ for each sub-term~$t_i$; apply $\Rule{Op}$ to all the derivations.
\end{itemize}
\end{proof}

@@

\begin{corollary}[Mode preprocessing]\label{thm:mode-decoration}
  It is decidable whether $V |-_{\Sigma,\Omega} \isTerm{t}^\dir{d}$.%for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$
\end{corollary}

@@

\begin{proof}[of \cref{thm:annotatability}]
By \cref{thm:generalised-mode-decoration}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.
Perform induction on this generalised mode derivation to construct a bidirectional typing derivation in the same mode (and also the term~$t'$ and a proof of $t' \sqsupseteq t$, which are determined by the derivations and omitted here).
\begin{itemize}
\item $\SynRule{Var}$, $\SynRule{Anno}$, and $\Rule{Op}$:
The outermost rule of the given typing derivation must be the corresponding one; apply the corresponding bidirectional typing rule to the induction hypotheses.
\item $\ChkRule{Sub}$:
Apply $\ChkRule{Sub}$ to the induction hypothesis.
\item $\SynRule{Missing}$:
This is the interesting case where we map a $\SynRule{Missing}$ rule to an $\SynRule{Anno}$ rule and add to the term a type annotation, which comes from the given typing derivation.
\end{itemize}
\end{proof}
